---
title: "Chapter 2: Data and measurement"
author: "A Solomon Kurz"
date: "`r format(Sys.Date())`"
output:
  github_document
---

```{r, eval = F, echo = F}
https://github.com/avehtari/ROS-Examples/
```

```{r, echo = F, cache = F}
knitr::opts_chunk$set(fig.retina = 2.5)
knitr::opts_chunk$set(fig.align = "center")
# options(width = 100)
```

# Data and measurement

> In this book, we'll be fitting lines (and some curves) to data, making comparisons and predictions and assessing our uncertainties in the resulting inferences. We'll discuss the assumptions underlying regression models, methods for checking these assumptions, and directions for improving fitted models. We'll discuss the challenges of extrapolating from available data to make causal inferences and predictions for new data, and we’ll use computer simulations to summarize the uncertainties in our estimates and predictions.
>
> Before fitting a model, though, it is a good idea to understand where your numbers are coming from. (p. 21)

## 2.1 Examining where data come from

Load the `hdi.dat` and `votes.dta` data files.

```{r, message = F, warning = F}
library(tidyverse)

# hdi <- read_table2("ROS-Examples-master/HDI/data/hdi.dat", col_names = T)
hdi <- read.table("ROS-Examples-master/HDI/data/hdi.dat", header = T)
# votes <- haven::read_dta("ROS-Examples-master/HDI/data/state vote and income, 68-00.dta")
votes <- foreign::read.dta("ROS-Examples-master/HDI/data/state vote and income, 68-00.dta")

glimpse(hdi)
glimpse(votes)
```

The two data sets can be joined by State, but only `hdi` contains a row for `Washington, D.C.`.

```{r}
anti_join(
  hdi %>% distinct(state),
  votes %>% distinct(st_state) %>% rename(state = st_state),
  by = "state"
)
```

Before we plot, let's set our default theme settings.

```{r}
theme_set(theme_linedraw() +
            theme(panel.grid = element_blank()))
```

For Figure 2.1, we'll need help from the **urbnmapr** package.

```{r}
library(urbnmapr)

left_join(
  get_urbn_map(map = "states", sf = TRUE),
  left_join(hdi,
            votes %>% mutate(state = st_state),
            by = "state") %>% 
    mutate(state_name = ifelse(state == "Washington, D.C.", "District of Columbia", state)),
  by = "state_name"
  ) %>% 
  # discretize the HDI variable
  mutate(hdi = case_when(
    hdi >= .95 ~ ".950+",
    hdi >= .90 ~ ".900-.949",
    hdi >= .85 ~ ".850-.899",
    hdi >= .80 ~ ".800-.849",
    hdi >= .75 ~ ".750-.799"
  )) %>%
  # now make hdi a factor
  mutate(hdi = factor(hdi,
                      levels = c(".950+", ".900-.949", ".850-.899", ".800-.849", ".750-.799"))) %>% 
  
  # plot!
  ggplot() +
  geom_sf(aes(fill = hdi, geometry = geometry),
          size = .1, color = "white") +
  scale_fill_viridis_d("Human Development Index by State",
                       option = "E",
                       guide = guide_legend(
                         direction = "horizontal",
                         keywidth = unit(1.5, "cm"),
                         label.hjust = 0,
                         label.position = "bottom",
                         title.position = "top")
                       ) + 
  theme_void() +
  theme(legend.position = "bottom")
```

Figure 2.2 shows the HDI is largely a function of state income.

```{r, fig.width = 8, fig.height = 3.75}
# left
p1 <-
  left_join(hdi,
            votes %>% mutate(state = st_state),
            by = "state") %>% 
  filter(st_year == 2000) %>% 
  
  ggplot(aes(x = st_income, y = hdi, label = st_stateabb)) +
  geom_text() +
  labs(x = "Average state income in 2000", 
       y = "Human Development Index")

# right
p2 <-
  left_join(hdi,
            votes %>% mutate(state = st_state),
            by = "state") %>% 
  filter(st_year == 2000) %>% 
  mutate(income_rank = min_rank(st_income),
         hdi_rank = min_rank(hdi)) %>% 
  
  ggplot(aes(x = income_rank, y = hdi_rank, label = st_stateabb)) +
  geom_text() +
  labs(x = "Rank of average state income in 2000", 
       y = "Rank of Human Development Index")

# combine
library(patchwork)

p1 + p2
```

### 2.1.1 Details of measurement can be important.

Load the `pew_research_center_june_elect_wknd_data.dta` data.

```{r}
# pew_pre <- haven::read_dta("ROS-Examples-master/Pew/data/pew_research_center_june_elect_wknd_data.dta")
pew_pre <- foreign::read.dta("ROS-Examples-master/Pew/data/pew_research_center_june_elect_wknd_data.dta")

glimpse(pew_pre)
```

The panels in Figure 2.3 take a bit of pre-plotting wrangling. Here's the code for Figure 2.3.a.

```{r}
# wrangle and save
d <-
  pew_pre %>% 
  drop_na(ideo, income) %>% 
  filter(ideo != "dk/refused") %>% 
  filter(income != "dk/refused") %>% 
  group_by(income) %>% 
  count(ideo) %>% 
  mutate(percent = 100 * n / sum(n)) %>% 
  mutate(income = case_when(
    income == "less than $10,000" ~ 1,
    income == "$10,000-$19,999" ~ 2,
    income == "$20,000-$29,999" ~ 3,
    income == "$30,000-$39,999" ~ 4,
    income == "$40,000-$49,000" ~ 5,
    income == "$50,000-$74,999" ~ 6,
    income == "$75,000-$99,999" ~ 7,
    income == "$100,000-$149,999" ~ 8,
    income == "$150,000+" ~ 9
  )) 

p1 <-
  d %>% 
  ggplot(aes(x = income, y = percent)) +
  geom_area(aes(fill = ideo)) +
  # we need further wrangling to make the label coordinates
  geom_text(data = d %>% 
              filter(income == 5) %>% 
              arrange(desc(ideo)) %>% 
              mutate(percent = cumsum(percent) - percent / 2),
            aes(label = ideo, color = percent < 30),
            size = 3) +
  scale_fill_viridis_d(direction = -1, breaks = NULL) +
  scale_color_manual(values = c("black", "white"), breaks = NULL) +
  scale_x_continuous(NULL, breaks = 1:9,
                     labels = c("", "Low income", "", "", "Middle income", "", "", "High income", ""),
                     expand = c(0, 0)) +
  scale_y_continuous(NULL, breaks = 0:2 * 50, labels = function(x) str_c(x, "%"),
                     expand = c(0, 0)) +
  labs(subtitle = "Self−declared political ideology, by income")
```

Now make Figure 2.3.b.

```{r}
# wrangle and save
d <-
  pew_pre %>% 
  drop_na(party) %>% 
  mutate(pid = case_when(
    party == "republican" ~ "Republican",
    party == "democrat" ~ "Democrat",
    partyln == "lean republican" ~ "Lean Rep.",
    partyln == "lean democrat" ~ "Lean Dem.",
    # party == "independent" ~ "Independent"
    TRUE ~ "Independent"
  ),
  income = case_when(
    income == "less than $10,000" ~ 1,
    income == "$10,000-$19,999" ~ 2,
    income == "$20,000-$29,999" ~ 3,
    income == "$30,000-$39,999" ~ 4,
    income == "$40,000-$49,000" ~ 5,
    income == "$50,000-$74,999" ~ 6,
    income == "$75,000-$99,999" ~ 7,
    income == "$100,000-$149,999" ~ 8,
    income == "$150,000+" ~ 9
  )) %>% 
  mutate(party = factor(pid,
                        levels = c("Republican", "Lean Rep.", "Independent", "Lean Dem.", "Democrat"))) %>% 
  drop_na(income) %>% 
  group_by(income) %>% 
  count(party) %>% 
  mutate(percent = 100 * n / sum(n))
 
p2 <-
  d %>% 
  ggplot(aes(x = income, y = percent)) +
  geom_area(aes(fill = party)) +
  # we need further wrangling to make the label coordinates
  geom_text(data = d %>%
              filter(income == 5) %>%
              arrange(desc(party)) %>%
              mutate(percent = cumsum(percent) - percent / 2),
            aes(label = party, color = percent < 50),
            size = 3) +
  scale_fill_viridis_d(direction = -1, breaks = NULL) +
  scale_color_manual(values = c("black", "white"), breaks = NULL) +
  scale_x_continuous(NULL, breaks = 1:9,
                     labels = c("", "Low income", "", "", "Middle income", "", "", "High income", ""),
                     expand = c(0, 0)) +
  scale_y_continuous(NULL, breaks = 0:2 * 50, labels = function(x) str_c(x, "%"),
                     expand = c(0, 0)) +
  labs(subtitle = "Self−declared party identification, by income")
```

Combine the two panels and finally make Figure 2.3.

```{r, fig.width = 8, fig.height = 3.25}
p1 + p2
```

If you look closely, you'll see our plot on the right is a little different than the one in the text. The difference seems to be in how we partitioned off the `"Independent"` and `"Lean Dem."` categories. For closer look, compare our code with Vehtari's code at [https://github.com/avehtari/ROS-Examples/blob/master/Pew/pew.Rmd](https://github.com/avehtari/ROS-Examples/blob/master/Pew/pew.Rmd).

## 2.2 Validity and reliability

### 2.2.1 Validity.

"We can define the *validity* of a measuring process as the property of giving the right answer on average across a wide range of plausible scenarios" (pp. 24--25, *emphasis* in the original).

### 2.2.2 Reliability.

"A *reliable* measure is one that is precise and stable" (p. 25, *emphasis* in the original).

### 2.2.3 Sample selection.

This stems from "the idea that the data you see can be a nonrepresentative sample of a larger population that you will not see" (p. 25).

## 2.3 All graphs are comparisons

"We can learn a lot by looking at data with an open mind. We present three quick examples here" (p. 25).

### 2.3.1 Simple scatterplots.

Load the `healthdata.txt`.

```{r}
# health <- read_table2("ROS-Examples-master/HealthExpenditure/data/healthdata.txt")
health <- read.table("ROS-Examples-master/HealthExpenditure/data/healthdata.txt", header = T)

glimpse(health)
```

Here's how to make Figure 2.4.

```{r, fig.width = 4.5, fig.height = 3}
remove <-  c("Netherlands", "Belgium", "Germany", "Ireland", "Iceland", "Greece", "Italy", "Sweden", "Finland")


health %>%
  filter(!country %in% remove) %>% 
  
  ggplot(aes(x = spending, y = lifespan, label = country)) +
  geom_text(size = 3) +
  scale_x_continuous("Health care spending (PPP US$)",
                     expand = expansion(mult = c(0, 0.05)), limits = c(0, NA)) +
  scale_y_continuous("Life expectancy (years)", breaks = 0:4 * 2 + 74)
```

### 2.3.2 Displaying more information on a graph.

It's not immediately clear which data they used to make Figure 2.5. Keep an eye out.

### 2.3.3 Multiple plots.

"We can learn by putting multiple related graphs in a single display" (p. 27). For this section, we need to load the `SSA-longtail-names.csv` data.

```{r, message = F, warning = F}
allnames <- 
  read_csv("ROS-Examples-master/Names/data/SSA-longtail-names.csv") %>% 
  mutate(sex = case_when(sex == FALSE ~ "F", 
                         is.na(sex) ~ "M"))
```

Make Figure 2.6.

```{r, message = F, fig.width = 4.25, fig.height = 2.75}
allnames %>% 
  filter(sex == "M") %>% 
  mutate(last = str_sub(name, -1)) %>% 
  group_by(last) %>% 
  summarise(sum = sum(`1906`)) %>% 
  ungroup() %>% 
  mutate(percent = 100 * sum / sum(sum)) %>% 
  
  ggplot(aes(x = last)) +
  geom_col(aes(y = percent)) +
  geom_text(aes(y = percent + 0.4, label = round(percent, digits = 1)),
            size = 2) +
  scale_y_continuous("Percentage of boys born", breaks = 0:1 * 10, 
                     expand = expansion(mult = c(0, 0.05))) +
  labs(subtitle = "Last letter of boys' names in 1906",
       x = NULL) +
  theme(axis.ticks.x = element_blank())
```

Here were the top ten most common boys' names in 1906.

```{r}
allnames %>% 
  filter(sex == "M") %>% 
  arrange(desc(`1906`)) %>% 
  rename(count = `1906`) %>% 
  select(name, count) %>% 
  slice(1:10)
```

Now make Figure 2.7.

```{r, message = F, fig.width = 8, fig.height = 2.75}
# 1956
p1 <-
  allnames %>% 
  filter(sex == "M") %>% 
  mutate(last = str_sub(name, -1)) %>% 
  group_by(last) %>% 
  summarise(sum = sum(`1956`)) %>% 
  ungroup() %>% 
  mutate(percent = 100 * sum / sum(sum)) %>% 
  
  ggplot(aes(x = last)) +
  geom_col(aes(y = percent)) +
  scale_y_continuous("Percentage of boys born", breaks = 0:1 * 10, 
                     expand = expansion(mult = c(0, 0.05))) +
  labs(subtitle = "Last letter of boys' names in 1956",
       x = NULL) +
  theme(axis.ticks.x = element_blank())

# 2006
p2 <-
  allnames %>% 
  filter(sex == "M") %>% 
  mutate(last = str_sub(name, -1)) %>% 
  group_by(last) %>% 
  summarise(sum = sum(`2006`)) %>% 
  ungroup() %>% 
  mutate(percent = 100 * sum / sum(sum)) %>% 
  
  ggplot(aes(x = last)) +
  geom_col(aes(y = percent)) +
  scale_y_continuous("Percentage of boys born", expand = expansion(mult = c(0, 0.05))) +
  labs(subtitle = "Last letter of boys' names in 2006",
       x = NULL) +
  theme(axis.ticks.x = element_blank())

# combine
p1 + p2
```

Here were the top ten most common boys' names ending in "n" in 2006.

```{r}
allnames %>% 
  filter(sex == "M") %>% 
  mutate(last  = str_sub(name, -1),
         count = `2006`) %>% 
  filter(last == "n") %>% 
  arrange(desc(count)) %>% 
  select(name, count) %>% 
  slice(1:10)
```

Figure 2.8.

```{r, message = F, fig.width = 5, fig.height = 3.25}
allnames %>% 
  filter(sex == "M") %>% 
  select(-X1, -sex) %>% 
  # select(name:`1950`) %>% 
  pivot_longer(-name,
               # `1880`:`1900`,
               names_to = "year",
               values_to = "count") %>% 
  mutate(last = str_sub(name, -1)) %>% 
  group_by(last, year) %>% 
  summarise(n = sum(count)) %>% 
  ungroup() %>%
  group_by(year) %>%
  mutate(percent = 100 * n / sum(n),
         year    = as.double(year),
         last    = toupper(last),
         size    = ifelse(last %in% c("D", "N", "Y"), "a", "b"),
         type    = case_when(
           last == "D" ~ "4",
           last == "N" ~ "1",
           last == "Y" ~ "2",
           TRUE ~ "1"
         )) %>% 
  
  ggplot(aes(x = year, y = percent, group = last)) +
  geom_line(aes(size = size, color = size, linetype = type)) +
  annotate(geom = "text",
           x = c(1938, 2001, 1977),
           y = c(19, 35, 14.5),
           label = c("D", "N", "Y")) +
  scale_color_manual(values = c("black", "grey50")) +
  scale_size_manual(values = c(0.6, 0.2)) +
  scale_x_continuous(expand = c(0, 0), breaks = 0:2 * 50 + 1900) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), limits = c(0, 40),
                     breaks = 0:2 * 20, labels = function(x) str_c(x, "%")) +
  labs(subtitle = "Last letters of boys' names",
       x = NULL,
       y = "Percentage of all boys' names that year") +
  theme(legend.position = "none")
```

Now make Figure 2.9

```{r, message = F, fig.width = 5, fig.height = 3.25}
# reshape
allnames %>% 
  select(-X1) %>% 
  pivot_longer(-c(name, sex),
               names_to = "year",
               values_to = "count") %>% 
  mutate(year = as.double(year)) %>% 
  # rank by sex and year
  group_by(year, sex) %>%
  mutate(rank = desc(count) %>% row_number()) %>% 
  # isolate the counts for the top 10 in each, making the other counts zeros
  mutate(count_top_10 = ifelse(rank <= 10, count, 0)) %>% 
  # compute the percentages
  summarise(percent = 100 * sum(count_top_10) / sum(count)) %>% 
  
  # plot!
  ggplot(aes(x = year, y = percent, group = sex)) +
  geom_line(size = 1/3) +
  annotate(geom = "text",
           x = c(1906, 1930),
           y = c(36, 20),
           label = c("Boys", "Girls")) +
  scale_x_continuous(expand = c(0, 0), breaks = 0:2 * 50 + 1900) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), limits = c(0, 45),
                     breaks = 0:2 * 20, labels = function(x) str_c(x, "%")) +
  labs(subtitle = "Total popularity of top 10 names each year, by sex",
       x = "Year",
       y = NULL) +
  theme(legend.position = "none")
```

Here we ranked the names using the `dplyr::row_number()` function. For a very similar but slightly different result, try ranking using `dplyr::min_rank()`. The two functions differ in how they deal with ties.

### 2.3.4 Grids of plots.

> A scatterplot displays two continuous variables, say $y$ vs. $x_1$. Coloring the dots enables us to plot a third variable, $x_2$, with some small number of discrete levels. Realistically it can be difficult to read a plot with more than two colors. We can then include two more discrete variables by constructing a two-way grid of plots representing discrete variables $x_3$ and $x_4$. This approach of *small multiples* can be more effective than trying to cram five variables onto a single plot. (p. 28, *emphasis* in the original)

To make Figure 2.10, we'll need the data in the `ROS-Examples-master/Congress/data` folder. Rather than living in a single file, the data are divided up by year in a series of `.asc` files. Here we'll follow some of the essence of Vehtari's code from [https://avehtari.github.io/ROS-Examples/Congress/congress_plots.html](https://avehtari.github.io/ROS-Examples/Congress/congress_plots.html) and load the data files into a list.

```{r, message = F, warning = F}
congress <-
  list.files(path = "ROS-Examples-master/Congress/data", full.names = TRUE)[1:49] %>% 
  map(read.table)

# not shown
# str(congress)
```


```{r, eval = F, echo = F}
# you can pull the data like this, but the resulting nested tibble format
# isn't the easiest for shaping it into the format necessary to make the figure

tibble(file_path = list.files(path = "ROS-Examples-master/Congress/data", full.names = TRUE)) %>% 
  slice(1:49) %>% map(file_path, read_table2)
  mutate(d = map(file_path, read_table2))
```

The 49 data files now live in a list called `congress`. Here we'll write a custom `make_data()` function which will pull and arrange the necessary data for the figure.

```{r make_data}
make_data <- function(i) {

  cong1 <- congress[[i]]
  cong2 <- congress[[i + 1]]
      
  d <-
    tibble(state_code = cong1[, 1]) %>% 
    mutate(region  = floor(state_code / 20) + 1,
           inc     = cong1[, 3],
           dvote1  = cong1[, 4] / (cong1[, 4] + cong1[, 5]),
           dvote2  = cong2[, 4] / (cong2[, 4] + cong2[, 5])) %>% 
    mutate(contested = (abs(dvote1 - 0.5)) < 0.3 & (abs(dvote2 - 0.5) < 0.3))
  
  return(d)
  
}
```

Our custom `make_data()` function works like this.

```{r}
make_data(27)
```

Now we can finally make Figure 2.10.

```{r}
# define the target years
tibble(i = c(27, 37, 47)) %>% 
  mutate(year = 1896 + 2 * (i - 1)) %>% 
  # extract the relevant data
  mutate(d = map(i, make_data)) %>% 
  unnest(d) %>% 
  # filter
  filter(contested == T) %>% 
  filter(abs(inc) %in% 0:1) %>%
  # wrangle
  mutate(label = str_c(year, "\nto\n", year + 2)) %>% 
  mutate(region = factor(region,
                         levels = 1:4,
                         labels = c("Northeast", "Midwest", "South", "West"))) %>% 
  mutate(inc = abs(inc) %>% as.character()) %>% 
  # filter one last time
  filter(!is.na(region)) %>% 
  
  # plot!
  ggplot(aes(x = dvote1, y = dvote2 - dvote1)) +
  geom_hline(yintercept = 0, size = 1/4) +
  geom_point(aes(color = inc),
             size = 1/4) +
  scale_color_manual(values = c("black", "grey60"), breaks = NULL) +
  scale_x_continuous(breaks = c(.25, .5, .75), labels = c("25%", "50%", "75%")) +
  scale_y_continuous(breaks = c(-0.25, 0, 0.25), labels = c("-25%", "0", "25%"), limits = c(-.33, .33)) +
  labs(title = "Swings in U.S. congressional elections in three different periods.",
       subtitle = "Gray dots are for incumbents running for reelection, black are for open seats.",
       x = "Dem. vote in election 1",
       y = "Vote swing") +
  theme(strip.text.y = element_text(angle = 0)) +
  facet_grid(label~region)
```

### 2.3.5 Applying graphical principles to numerical displays and communication more generally.

"When reporting data and analysis, you should always imagine yourself in the position of the reader of the report. Avoid overwhelming the reader with irrelevant material" (p. 29).

If you would like to prevent **R** from providing too many significant digits, you can always use the `digits` argument within the `options()` function. For example, if you wanted to change the default settings to two significant digits, you might execute `options(digits = 2)`. As we will see, several of the convenience functions within the **brms** package have their own `digits` argument, which works the same way for local output.

> Never display a graph you can't talk about. Give a full caption for every graph, as we try to do in this book. This explains, to yourself and others, what you are trying to show and what you have learned from each plot. Avoid displaying graphs that have been made simply because they are conventional. (p. 30)

### 2.3.6 Graphics for understanding statistical models.

The authors suggest we can summarize the kinds of graphs we'll be making in this text as falling into three kinds:

* displays of raw data;
* plots of fitted models and inferences; and
* plots presenting the final results, used as communication tools.

### 2.3.7 Graphs as comparisons.

> All graphical displays can be considered as comparisons. When making a graph, line things up so that the most important comparisons are clearest. Comparisons are clearest when scales are lined up. (p. 31)

### 2.3.8 Graphs of fitted models.

"It can be helpful to graph a fitted model and data on the same plot" (p. 31).

## 2.4 Data and adjustment: trends in mortality rates

"Even when there are no questions of data quality or modeling, it can make sense to adjust measurements to answer real-world questions" (p. 31). To see why, load the `white_nonhisp_death_rates_from_1999_to_2013_by_sex.txt` data.

```{r, warning = F, message = F}
death <- read_table2("ROS-Examples-master/AgePeriodCohort/data/white_nonhisp_death_rates_from_1999_to_2013_by_sex.txt")

glimpse(death)
```

Make Figure 2.11a

```{r, message = F, fig.width = 4.25, fig.height = 2.875}
p1 <-
  death %>% 
  filter(Age >= 45 & Age <= 54) %>% 
  group_by(Year) %>% 
  summarise(Deaths     = sum(Deaths),
            Population = sum(Population)) %>% 
  mutate(dr = Deaths / Population) %>% 
  
  ggplot(aes(x = Year, y = dr)) +
  geom_line() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(subtitle = "Raw death rates\nfor 45−54−year−old non−Hisp whites",
       x = NULL,
       y = "Death rate among non−Hisp whites 45−54") +
  theme(axis.text = element_text(size = 9),
        axis.title = element_text(size = 9))

p1
```

Figure 2.11b.

```{r, message = F, fig.width = 4.25, fig.height = 2.875}
p2 <-
  death %>% 
  filter(Age >= 45 & Age <= 54) %>% 
  group_by(Year) %>% 
  summarise(mu_age = sum(Age * Population) / sum(Population)) %>% 
  
  ggplot(aes(x = Year, y = mu_age)) +
  geom_line() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(subtitle = "But the average age in this group is going up!",
       x = NULL,
       y = "Avg age among non−Hisp whites 45−54") +
  theme(axis.text = element_text(size = 9),
        axis.title = element_text(size = 9))

p2
```

> Suppose for the moment that mortality rates did not change for individuals in this age group from 1999 to 2013. In this case, we could calculate the change in the group mortality rate due solely to the change in the underlying age of the population. We do this by taking the 2013 mortality rates for each age and computing a weighted average rate each year using the number of individuals in each age group. (p. 32)

We need a little preparatory work before making Figure 2.11c. Here we compute and save the death rates for 2013.

```{r, message = F}
 rate_2013 <-
  death %>% 
  # select our focal `Age` groups
  filter(Age >= 45 & Age <= 54) %>% 
  # use only 2013, which will define our standard population
  filter(Year == 2013) %>% 
  # collapse across `Male`
  group_by(Age) %>% 
  summarise(rate_2013 = sum(Deaths) / sum(Population)) %>% 
  ungroup()

rate_2013
```

Now we make Figure 2.11.c by combining the crude death rate and the adjusted rate, which is based on what it would look like if the age-specific mortality rates had been at the 2013 level for all the years.

```{r, message = F, fig.width = 10, fig.height = 2.5}
p3 <-
  left_join(
    # crude_rate
    death %>% 
      filter(Age >= 45 & Age <= 54) %>% 
      group_by(Year) %>% 
      summarise(crude_rate = sum(Deaths) / sum(Population)) ,
    
    # adjusted_rate
    death %>% 
      # select our focal `Age` groups
      filter(Age >= 45 & Age <= 54) %>% 
      # collapse across `Male`
      group_by(Year, Age) %>% 
      summarise(Deaths = sum(Deaths),
                Population = sum(Population)) %>% 
      group_by(Year) %>% 
      mutate(proportion = Population / sum(Population)) %>% 
      left_join(rate_2013, by = "Age") %>% 
      summarise(adjusted_rate = sum(proportion * rate_2013)),
  ) %>% 
  
  # make the plot
  ggplot(aes(x = Year)) +
  geom_line(aes(y = crude_rate),
            color = "grey50", size = 1/3) +
  geom_line(aes(y = adjusted_rate)) +
  annotate(geom = "text",
           x = c(2000.5, 1999.33), y = c(.0039, .00409),
           label = c("Raw death rate", "Expected just from\nage shift"),
           color = c("grey40", "black"), hjust = 0, size = 3) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(subtitle = "The trend in raw death rates since 2005\ncan be explained by age−aggregation bias",
       x = NULL,
       y = "Death rate for 45−54 non−Hisp whites")

# combine all three panels and plot!
(p1 + p2 + p3) &
  theme(axis.text = element_text(size = 7),
        axis.title = element_text(size = 7),
        title = element_text(size = 9))
```

> Contrary to the original claim from the raw numbers, we find there is no longer a steady increase in mortality rates for this age group after adjusting for age composition. Instead, there is an increasing trend from 1999 to 2005 and a constant trend thereafter. (p. 31)

Figure 2.12 takes a complimentary approach. 

> Having demonstrated the importance of age adjustment, we now perform an adjustment for the changing age composition. We ask what the data would look like if the age groups remained the same each year and only the individual mortality rates changed. Figure 2.12a shows the simplest such adjustment, normalizing each year to a hypothetical uniformly distributed population in which the number of people is equal at each age from 45 through 54. That is, we calculate the mortality rate each year by dividing the number of deaths for each age between 45 and 54 by the population of that age and then taking the average. This allows us to compare mortality rates across years. (p. 32)

Here's our Figure 2.12a.

```{r, message = F, fig.width = 4.25, fig.height = 2.875}
adjusted_for_age_comp <-
  death %>% 
  # select our focal `Age` groups
  filter(Age >= 45 & Age <= 54) %>% 
  group_by(Year, Age) %>% 
  summarise(rate_by_year_and_age = sum(Deaths) / sum(Population)) %>% 
  group_by(Year) %>% 
  summarise(rate_by_year_and_average_age = mean(rate_by_year_and_age))

# save the `rate_by_year_and_average_age` value for 1999
adjusted_for_age_comp_1999 <-
  adjusted_for_age_comp %>% 
  filter(Year == 1999) %>% 
  pull(rate_by_year_and_average_age)

# normalize
p1 <-
  adjusted_for_age_comp %>% 
  mutate(normalized_rates = rate_by_year_and_average_age / adjusted_for_age_comp_1999) %>% 
  
  # plot
  ggplot(aes(x = Year, y = normalized_rates)) +
  geom_line() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(subtitle = "Trend in age−adjusted death rate\nfor 45−54−year−old non−Hisp whites",
       x = NULL,
       y = "Age−adj death rate, relative to 1999") +
  theme(axis.text = element_text(size = 9),
        axis.title = element_text(size = 9))

p1
```

`r emo::ji("warning")` The rest of this section needs a little work. I'll come back to this, later. `r emo::ji("warning")`

```{r, message = F}
proportion <-
  death %>% 
  # select our focal `Age` groups
  filter(Age >= 45 & Age <= 54) %>% 
  # collapse across `Male`
  group_by(Year, Age) %>% 
  summarise(Population = sum(Population)) %>% 
  group_by(Year) %>% 
  mutate(proportion = Population / sum(Population)) %>% 
  ungroup()

proportion
```

```{r, message = F}
death %>% 
  # select our focal `Age` groups
  filter(Age >= 45 & Age <= 54) %>% 
  group_by(Year, Age) %>% 
  summarise(rate_by_year_and_age = sum(Deaths) / sum(Population)) %>% 
  left_join(
    proportion %>% 
      filter(Year == 1999) %>% 
      select(Age, Population),
    by = "Age"
  ) %>% 
  group_by(Year) %>% 
  summarise(rate_by_year_and_average_age = mean(rate_by_year_and_age),
            rate_by_year_and_average_age_1999 = mean(Population * rate_by_year_and_age))
```

```{r, message = F}
adjusted_for_age_comp %>% 
  
  mutate(normalized_rates = rate_by_year_and_average_age / adjusted_for_age_comp_1999)


  
  # plot
  # ggplot(aes(x = Year, y = normalized_rates)) +
  # geom_line() +
  # scale_x_continuous(expand = c(0, 0)) +
  # labs(subtitle = "Trend in age−adjusted death rate\nfor 45−54−year−old non−Hisp whites",
  #      x = NULL,
  #      y = "Age−adj death rate, relative to 1999") +
  # theme(axis.text = element_text(size = 9),
  #       axis.title = element_text(size = 9))
```


```{r, eval = F, echo = F}
standard_populaiton <-
  death %>% 
  # select our focal `Age` groups
  filter(Age >= 45 & Age <= 54) %>% 
  # use only 2013, which will define our standard population
  filter(Year == 2013) %>% 
  # collapse across `Male`
  group_by(Age) %>% 
  summarise(standard_populaiton = sum(Population)) %>% 
  ungroup()

death %>% 
  # select our focal `Age` groups
  filter(Age >= 45 & Age <= 54) %>% 
  # collapse across `Male`
  group_by(Age, Year) %>% 
  summarise(Deaths = sum(Deaths),
            Population = sum(Population)) %>% 
  mutate(crude_rate = Deaths / Population) %>% 
  left_join(standard_populaiton,
            by = "Age") %>% 
  mutate(expected_deaths = crude_rate * standard_populaiton) %>% 
  # proportion based on the standard population (2013)
  group_by(Year) %>% 
  mutate(proportion = standard_populaiton / sum(standard_populaiton)) %>% 
  mutate(age_specific_adjusted_rate = crude_rate * proportion) %>% 
  summarise(age_adjusted_rate = sum(age_specific_adjusted_rate)) %>% 
  
  ggplot(aes(x = Year)) +
  geom_line(aes(y = age_adjusted_rate) )+
  geom_line(aes(y = age_adjusted_rate))
  
  
```




## Session info {-}

```{r}
sessionInfo()
```

```{r, warning = F, echo = F, eval = F}
rm(list = ls())
```

```{r, echo = F, message = F, warning = F, results = "hide", eval = F}
ggplot2::theme_set(ggplot2::theme_grey())
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

